---
title: "Repeating Using Loops"
keypoints:
- Use `for` to repeat operations.
- Use `while` to repeat operations until a stopping point.
objectives:
- Write and understand `for()` loops.
- Write and understand `while()` loops.
output: word_document
questions: How can I repeat operations in R?
source: Rmd
teaching: 30
exercises: 30
---

```{r, include=FALSE}
# Silently set seed for random number generation, so we don't have to explain it
set.seed(10)
```

## Repeating operations with a `for()` loop

If you want to iterate over
a set of values, when the order of iteration is important, and perform the
same operation on each, a `for()` loop will do the job.
Avoid using `for()` loops unless the order of iteration is important:
i.e. the calculation at each iteration depends on the results of previous iterations.

The basic structure of a `for()` loop is:

```{r, eval=FALSE}
for(iterator in set of values){
  do a thing
}
```

For example:

```{r}
for(i in 1:10){
  print(i)
}
```

The `1:10` bit creates a vector on the fly; you can iterate over any other vector as well.

We can use a `for()` loop nested within another `for()` loop to iterate over two things at
once.

```{r}
for(i in 1:5){
  for(j in c('a', 'b', 'c')){
    print(paste("We are on loop part", i,j))
  }
}
```

Rather than printing the results, we could write the loop output to a new object.

```{r}
output_vector <- c()
for(i in 1:5){
  for(j in c('a', 'b', 'c')){
    temp_output <- paste(i, j)
    output_vector <- c(output_vector, temp_output)
  }
}
output_vector
```

This approach can be useful, but 'growing your results' (building
the result object incrementally) is computationally inefficient, so avoid
it when you are iterating through a lot of values.

> ## Tip: don't grow your results
>
> One of the biggest things that trips up novices and
> experienced R users alike, is building a results object
> (vector, list, matrix, data frame) as your for loop progresses.
> Computers are very bad at handling this, so your calculations
> can very quickly slow to a crawl. It's much better to define
> an empty results object before hand of the appropriate dimensions.
> So if you know the end result will be stored in a matrix like above,
> create an empty matrix with 5 row and 5 columns, then at each iteration
> store the results in the appropriate location.


A better way is to define your (empty) output object before filling in the values.
For this example, it looks more involved, but is still more efficient.

```{r}
output_matrix <- matrix(nrow=3, ncol=3)
j_vector <- c('a', 'b', 'c')
for(i in 1:3){
  for(j in 1:3){
    temp_j_value <- j_vector[j]
    temp_output <- paste(i, temp_j_value)
    output_matrix[i, j] <- temp_output
  }
}
output_vector2 <- as.vector(output_matrix)
output_vector2
```

## Repeating operations until a condition is met

Sometimes you will find yourself needing to repeat an operation as long as a certain condition is met. You can do this with a `while()` loop.

```{r, eval=FALSE}
while(this condition is true){
   do a thing
  }
```

R will interpret a condition being met as "TRUE".
As an example, here's a while loop
that generates random numbers from a uniform distribution (the `runif()` function)
between 0 and 1 until it gets one that's less than 0.1.

```{r}
 z <- 1
while(z > 0.1){
   z <- runif(1)
   cat(z, "\n")
 }
```

`while()` loops will not always be appropriate. You have to be particularly careful that you don't end up stuck in an infinite loop because your condition is always met and hence the while statement never terminates.


> ## Challenge 1
>
> Write a script that loops through the `mtcars` data by cylinder number and prints out
> the mean mpg (miles per gallon) for each category of cyl (cylinders).
>
> Let's think about how to do this in steps.  Treat the answers to these questions as a worksheet leading up to step 6, where you put the answers together and do a little more more coding to create the final loop.

> 1. What are the variables (columns) we need?
> 2. Which combination of rows and columns are we getting the mean for?
> 3. How can we generate a mean for one group (rows containing 4 cylinders only, for example) for that column?
> 4. Which values need to be substituted, i.e, repeated, to get all the groups?
> 5. How do we automatically get that vector of unique values (group identifiers)?
> 6. Then and ONLY then, write the full loop.

> HINTS: Remember the subset notation dataframe[rows, columns]
> Remember functions you may need: unique(), mean(), cat(), names()

## Solution to Challenge 1

**Step 1** What are the variables (columns) we need?
```{r, eval=FALSE}
names(mtcars)
```

**Step 2** Which combination of rows and columns are we getting the mean for?
means of mpg for rows where cyl = 4, 6, and 8

**Step 3** How can we generate a mean for one group (rows containing 4 cylinders only, for example)?
```{r, eval=FALSE}
mtcars[mtcars$cyl == 4, "mpg"]
# then wrap the previous line of code in mean()
mean(mtcars[mtcars$cyl == 4, "mpg"])
```
**Step 4**: Which values need to be substituted, i.e, repeated, to get all the groups?
Per step 2, we need cylinder (mtcars$cyl) to have values of 4, 6, and 8, so replace the 4 with iCyl (an arbitrary but descriptive name).
```{r, eval=FALSE}
mean(mtcars[mtcars$cyl == iCyl, "mpg"])
```
**Step 5**: How do we automatically get that vector of unique values (group identifiers)?
```{r, eval=FALSE}
unique(mtcars$cyl) #cyl is the group identifier for each row
```
**Step 6**: Then and ONLY then, write the full loop.
Keep in mind for each iteration (repeat) of the loop, you can overwrite a previous value.
```{r, eval=FALSE}
for( iCyl in unique(mtcars$cyl) ){
#temporarily store the dataframe per group
  tmp <- mtcars[mtcars$cyl == iCyl, "mpg"] 
  cat(iCyl,
      "cylinders have",
      mean(tmp, na.rm = TRUE), 
      "mpg",
      sep = "\n")  
}
```


