---
title: "Repeating Using Loops"
keypoints:
- Use `for` to repeat operations.
- Use `while` to repeat operations until a stopping point.
objectives:
- Write and understand `for()` loops.
- Write and understand `while()` loops.
output: word_document
questions: How can I repeat operations in R?
source: Rmd
teaching: 30
exercises: 30
---

```{r, include=FALSE}
# Silently set seed for random number generation, so we don't have to explain it
set.seed(10)
```

## Repeating operations with a `for()` loop

If you want to iterate over
a set of values, when the order of iteration is important, and perform the
same operation on each, a `for()` loop will do the job.
Avoid using `for()` loops unless the order of iteration is important:
i.e. the calculation at each iteration depends on the results of previous iterations.

The basic structure of a `for()` loop is:

```{r, eval=FALSE}
for(iterator in set_of_values){
  do a thing
}
```

This great image shows iteration as a "monster parade". The for loop they enter has an if-else statement: if the monster is a triangle, it gets sunglasses. Otherwise, it gets a hat. The output is the parade of monsters where the same input parade of monsters shows up, now wearing either sunglasses (if triangular) or a hat (if any other shape).

![Illustrated for loop where the input vector is a parade of monsters, including monsters that are circles, triangles, and squares, as described by the text. Text and image CC-BY Allison Horst ](https://pbs.twimg.com/media/FrWkidtaQAIHC-5?format=jpg&name=4096x4096)


For example:

```{r}
parade <- c("Circle", "Square", "Triangle")
for(monster in parade){
  print(monster)
}
```

We can use a `for()` loop nested within another `for()` loop to iterate over two things at once, including the character vector represented here by j.

```{r}
for(monster in parade){
  for(letter in c('a', 'b', 'c')){
    print(paste("Monster named",
                monster, 
                "is learning about the letter",
                letter))
  }
}
```

Rather than directly calling the characters a through c, let's use their element indexing.  This will make it easier to understand how we can use this later in matrices or data frames.
```{r}
(alphabet <- c('a', 'b', 'c'))

for(monster in 1:3){
  for(letter in 1:3){
    print(paste("Monster named",
                parade[monster], 
                "is learning about the letter",
                alphabet[letter]))
  }
}
```

Next, rather than using the current final value of the vector length (here, 3), let's use a function to find it automatically in case we change the vector later.
```{r}
(alphabet <- c('a', 'b', 'c'))

for(monster in 1:length(parade)){
  for(letter in 1:length(alphabet)){
    print(paste("Monster named",
                parade[monster], 
                "is learning about the letter",
                alphabet[letter]))
  }
}
```

Next, let's also start storing the results instead of printing them.

```{r}
(alphabet <- c('a', 'b', 'c'))
(output_vector <- c())
for(monster in 1:length(parade)){
  for(letter in 1:length(alphabet)){
    temp_output <- paste("Monster named",
                         parade[monster], 
                         "is learning about the letter",
                         alphabet[letter])
    output_vector <- c(output_vector, temp_output)
  }
}
output_vector
```

This approach can be useful, but 'growing your results' (building
the result object incrementally) is computationally inefficient, so avoid
it when you are iterating through a lot of values.

> ## Tip: don't grow your results
>
> One of the biggest things that trips up novices and
> experienced R users alike, is building a results object
> (vector, list, matrix, data frame) as your for loop progresses.
> Computers are very bad at handling this, so your calculations
> can very quickly slow to a crawl. It's much better to define
> an empty results object before hand of the appropriate dimensions.
> So if you know the end result will be stored in a matrix like above,
> create an empty matrix with 3 rows and 3 columns, then at each iteration
> store the results in the appropriate location.


A better way is to define your (empty) output object before filling in the values.
For this example, it looks more involved, but is still more efficient.

```{r}
output_matrix <- matrix(nrow=3, ncol=3)
(alphabet <- c('a', 'b', 'c'))
for(monster in 1:length(parade)){
  for(letter in 1:length(alphabet)){
    temp_output <- paste("Monster named",
                         parade[monster], 
                         "is learning about the letter",
                         alphabet[letter])
    output_matrix[monster, letter] <- temp_output
  }
}
View(output_matrix)
```




> ## Challenge 1
>
> Write a script that loops through the `mtcars` data by column number and stores the mean for each column in a vector.
> Let's think about how to do this in steps.  Treat the answers to these questions as a worksheet leading up to step 6, where you put the answers together and do a little more more coding to create the final loop.

> 1. Write comments for the steps you will need in the loop.  Identify the storage vector, the iterator, the set of column numbers, and where you will perform your action (calculation of mean and storage of mean).
> 2. Create your 1:n (set of column numbers) by finding out what the column number is for the final column in mtcars. Use ncol (instead of length) to get that number for a data frame.
> 3. Create the code to get a single column's mean using subset notation. dataframe[row, column].
> 4. Replace the sample column number (here, column 1) with your iterator (here, cn)
> 5. Add in empty vector to prepare for storing all column means.
> 6. Store means by their column number to create a vector of means.  This lets can store all results, not just most recent overwrite of loop.
> 7. Add a print step AFTER the loop to view the final vector.

> HINTS: Remember the subset notation dataframe[rows, columns]
> Remember functions you may need: ncol(), mean()

## Solution to Challenge 1

Step 1. Write comments for the steps you will need in the loop.  Identify the storage vector, the iterator, the set of column numbers, and where you will perform your action (calculation of mean and storage of mean).
```{r}
# store results in a vector before the loop
# the loop itself
# for (cn in set_of_column_numbers) {
# calculate and store mean of column
# }
# view results
```

Step 2. What is the column number of the final column?  Take the code to get that number.  Replace "set_of_column_numbers" with the sequence of 1 through the final column number.

```{r}
ncol(mtcars)
# store results in a vector before the loop
# the loop itself
for (cn in 1:ncol(mtcars)) {
# calculate and store mean of column
}
# view results
```

Step 3. Create the code to get a single column using subset notation. dataframe[row, column].
```{r}
# store results in a vector before the loop
# the loop itself
for (cn in 1:ncol(mtcars)) {
# calculate mean and store mean of single column
col_means <- mean(mtcars[,1])
}
# view results
```
Step 4. Replace the sample column number (column 1) with your iterator (cn)
```{r}
# store results in a vector before the loop
# the loop itself
for (cn in 1:ncol(mtcars)) {
# calculate and store mean of column
col_means <- mean(mtcars[,cn])
}
# view results
```

Step 5. Add in empty vector to prepare for storing all column means.
```{r}
# store results in a vector before the loop
col_means <- c()
# the loop itself
for (cn in 1:ncol(mtcars)) {
# store mean of single column
col_means <- mean(mtcars[,cn])
}
# view results
```
Step 6. Store means by their column number to create a vector of means.  This lets can store all results, not just most recent overwrite of loop.
```{r}
# store results in a vector before the loop
col_means <- c()
# the loop itself
for (cn in 1:ncol(mtcars)) {
# store mean of single column
col_means[cn] <- mean(mtcars[,cn])
}
# view results
```

Step 7. Add a print step AFTER the loop to view the final vector.
```{r}
# store results in a vector before the loop
col_means <- c()
# the loop itself
for (cn in 1:ncol(mtcars)) {
# store mean of single column
col_means[cn] <- mean(mtcars[,cn])
}
# view results
print(col_means)
```