---
title: Vectorization
teaching: 10
exercises: 15
questions:
- "How can I operate on all the elements of a vector at once?"
objectives:
- "To understand vectorized operations in R."
keypoints:
- "Use vectorized operations instead of loops."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("09-")
# Silently load in the data so the rest of the lesson works
gapminder <- read.csv("data/gapminder_data.csv", header=TRUE)
library("ggplot2")
```

Most of R's functions are vectorized, meaning that the function will
operate on all elements of a vector without needing to loop through
and act on each element one at a time. This makes writing code more
concise, easy to read, and less error prone.


```{r}
x <- 1:4
x * 2
```

The multiplication happened to each element of the vector.

We can also add two vectors together:

```{r}
y <- 6:9
x + y
```

Each element of `x` was added to its corresponding element of `y`:

```{r, eval=FALSE}
x:  1  2  3  4
    +  +  +  +
y:  6  7  8  9
---------------
    7  9 11 13
```


> ## Challenge 1
>
> Let's try this on the `pop` column of the `gapminder` dataset.
>
> Make a new column in the `gapminder` data frame that
> contains population in units of millions of people.
> Check the head or tail of the data frame to make sure
> it worked.
>
> > ## Solution to challenge 1
> >
> > Let's try this on the `pop` column of the `gapminder` dataset.
> >
> > Make a new column in the `gapminder` data frame that
> > contains population in units of millions of people.
> > Check the head or tail of the data frame to make sure
> > it worked.
> >
> > ```{r}
> > gapminder$pop_millions <- gapminder$pop / 1e6
> > head(gapminder)
> > ```
> {: .solution}
{: .challenge}


> ## Challenge 2
>
> On a single graph, plot population, in
> millions, against year, for all countries. Don't worry about
>identifying which country is which.
>
> Repeat the exercise, graphing only for China, India, and
>Indonesia. Again, don't worry about which is which.
>
> > ## Solution to challenge 2
> >
> > Refresh your plotting skills by plotting population in millions against year.
> >
> > ```{r ch2-sol}
> > ggplot(gapminder, aes(x = year, y = pop_millions)) +
> >  geom_point()
> > countryset <- c("China","India","Indonesia")
> > ggplot(gapminder[gapminder$country %in% countryset,],
> >        aes(x = year, y = pop_millions)) +
> >   geom_point()
> > ```
> {: .solution}
{: .challenge}


Comparison operators, logical operators, and many functions are also
vectorized:


**Comparison operators**

```{r}
x > 2
```

**Logical operators**
```{r}
a <- x > 3  # or, for clarity, a <- (x > 3)
a
```

> ## Tip: some useful functions for logical vectors
>
> `any()` will return `TRUE` if *any* element of a vector is `TRUE`.  
> `all()` will return `TRUE` if *all* elements of a vector are `TRUE`.
{: .callout}

Most functions also operate element-wise on vectors:

**Functions**
```{r}
x <- 1:4
log(x)
```

Vectorized operations work element-wise on matrices:

```{r}
m <- matrix(1:12, nrow=3, ncol=4)
m * -1
```


> ## Tip: element-wise vs. matrix multiplication
>
> Very important: the operator `*` gives you element-wise multiplication!
> To do matrix multiplication, we need to use the `%*%` operator:
>
> ```{r}
> m %*% matrix(1, nrow=4, ncol=1)
> matrix(1:4, nrow=1) %*% matrix(1:4, ncol=1)
> ```
>
> For more on matrix algebra, see the [Quick-R reference
> guide](http://www.statmethods.net/advstats/matrix.html)
{: .callout}

