---
title: "Functions Explained"
keypoints:
- Use `function` to define a new function in R.
- Use parameters to pass values into functions.
- Load functions into programs using `source()`.
- Use `stopifnot()` to flexibly check function arguments in R.
objectives:
- Define a function that takes arguments.
- Return a value from a function.
- Set default values for function arguments.
- Check argument conditions with `stopifnot()` in functions.
output: word_document
questions: How can I write a new function in R?
source: Rmd
teaching: 45
exercises: 15
---
```{r setup, cache = F, include = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

If we only had one data set to analyze, it would probably be faster to load the
file into a spreadsheet and use that to plot simple statistics. However, 
our data may be updated periodically, and we may want to pull in that new
information later and re-run our analysis again. We may also obtain similar data
from a different source in the future.

In this lesson, we'll learn how to write a function so that we can repeat
several operations with a single command.

> ## What is a function?
>
> Functions gather a sequence of operations into a whole, preserving it for
> ongoing use. Functions provide:
>
> * a name we can remember and invoke it by
> * relief from the need to remember the individual operations
> * a defined set of inputs and expected outputs
> * rich connections to the larger programming environment
>
> As the basic building block of most programming languages, user-defined
> functions constitute "programming" as much as any single abstraction can. If
> you have written a function, you are a computer programmer.

## Defining a function

Let's open a new R script file in a `functions/` directory within our working directory.  We will name the script
functions-lesson.R.

Within that functions file, let's define a function `conversion()` that converts values into a new value 2.14 times the original value.

```{r}
conversion <- function(value) {
  dat <- value
  converted <- dat*2.14
  return(converted)
}
```

We define `conversion()` by assigning it to the output of `function`. The
list of argument names are contained within parentheses.   Next, the
body of the function--the
statements that are executed when it runs--is contained within curly braces
(`{}`). The statements in the body are indented by two spaces. This makes the
code easier to read but does not affect how the code operates.

It is useful to think of creating functions like writing a cookbook. First you define the "ingredients" that your function needs (these are the arguments). In this case, we only need one ingredient to use our function: "value". After we list our ingredients, we then say what we will do with them. In this case, we are taking our ingredient and applying a mathematical operators to it (multiplication). 

> ## Tip: Pass by value
>
> Functions in R almost always make copies of the data to operate on
> inside of a function body. When we modify `dat` inside the function
> we are modifying the copy of the mtcars dataset stored in `dat`,
> not the original variable we gave as the first argument.
>
> This is called "pass-by-value" and it makes writing code much safer:
> you can always be sure that whatever changes you make within the
> body of the function, stay inside the body of the function.


When we call the function, the values we pass to it as arguments are assigned to
those variables so that we can use them inside the function. Inside the
function, we use a return statement to send a result back to us.

> ## Tip
>
> One feature unique to R is that the return statement is not required.
> R automatically returns whichever variable is on the last line of the body
> of the function. But for clarity (and more complex returns), we will explicitly define the
> return statement.

## Loading functions from a separate file for cleaner code

As we mentioned before, we're going to write our functions into a separate R script. You can load in the functions into our R session by using the
`source()` function.  You can write functions directly into your original script file, but keeping functions cleanly separate is a good habit for reproducibility and ease of reading the code.

```{r, eval=FALSE}
source("functions/functions-lesson.R")
```


Let's try running our function.
Calling our own function is no different from calling any other function:

```{r}
# one inch to 2.14 cm
conversion(1)
```



> ## Challenge 1
>
> Write a function called `conversion_grams_to_kg()` that takes a mass in grams and returns the value in kilograms.
>
> Hint: To convert from grams to kilograms you divide by 1000
>
> > ## Solution to challenge 1
> >
> > ```{r}
> > conversion_grams_to_kg <- function(value) {
> >  dat <- value
> >  kg <- dat/1000
> >  return(kg)
> > }
> > 
> > conversion_grams_to_kg(1000)
> > ```

You can add more than one argument (or "ingredient") to the function.  In this case, let's make the conversion factor into an argument we can specify.
```{r}
conversion <- function(value, conversion_factor) {
  dat <- value
  converted <- dat*conversion_factor
  return(converted)
}
```
Test it out on two different values.
```{r}
# convert inches to cm
conversion(value = 1, conversion_factor = 2.14)
# convert grams to kilograms
conversion(value = 1000, conversion_factor = 0.0001)
```



> ## Tip: Function scope
>
> Another important concept is scoping: any variables (or functions!) you
> create or modify inside the body of a function only exist for the lifetime
> of the function's execution. When we call `conversion()`, the variable `converted` only exists inside the body of the function. Even if we
> have variables of the same name in our interactive R session, they are
> not modified in any way when executing a function.


## Defining Defaults

We can pass arguments to functions in two ways, directly or named.
Here's what it looks like to pass arguments without naming them.

```{r}
dat <- round(3.14, 2)
```

However, the position of the arguments matters if they are not named.

```{r, error = TRUE}
dat <- round(2, 3.14)
```
> ## Matching Arguments
>
> To be precise, R has three ways that arguments supplied
> by you are matched to the *formal arguments* of the function definition:
>
> 1. by complete name,
> 2. by partial name (matching on initial *n* characters of the argument name), and
> 3. by position.
>
> Arguments are matched in the manner outlined above in *that order*: by
> complete name, then by partial matching of names, and finally by position.

To understand what's going on, and make our own functions easier to use, let's re-define our `conversion` function like this:

```{r}
conversion <- function(value, conversion_factor=2.14) {
  dat <- value
  converted <- dat*conversion_factor
  return(converted)
}
```

The key change is that the second argument is now written `conversion_factor = 2.14` instead of just `conversion_factor`.
If we call the function with two arguments, it works as it did before:

```{r}
conversion(value = 1, conversion_factor = 10)
```

But we can also now call `conversion()` with just one argument, in which case `conversion_factor` is automatically assigned the default value of `2.14`:

```{r}
conversion(value = 1)
```

This is handy: if we usually want a function to work one way, but occasionally need it to do something else, we can allow people to pass an argument when they need to but provide a default to make the normal case easier.



## Defensive Programming

Now that we've begun to appreciate how writing functions provides an efficient
way to make R code re-usable and modular, we should note that it is important
to ensure that functions only work in their intended use-cases. Checking
function parameters is related to the concept of _defensive programming_.
Defensive programming encourages us to frequently check conditions and throw an
error if something is wrong. These checks are referred to as assertion
statements because we want to assert some condition is `TRUE` before proceeding.
They make it easier to debug because they give us a better idea of where the
errors originate.

### Checking conditions with `stopifnot()`

Let's start by re-examining `conversion()`, our function for converting
values. It was defined like so:

```{r}
conversion <- function(value, conversion_factor) {
  dat <- value
  converted <- dat*conversion_factor
  return(converted)
}
```

For this function to work as intended, the argument `value` must be a `numeric`
value; otherwise, the mathematical procedure for converting between the two
numeric units will not work. Luckily R provides the convenience function
`stopifnot()`. We can list as many requirements that should evaluate to `TRUE`;
`stopifnot()` throws an error if it finds one that is `FALSE`. Listing these
conditions also serves a secondary purpose as extra documentation for the
function.

Let's try out defensive programming with `stopifnot()` by adding assertions to
check the input to our function `conversion()`.

We want to assert the following: `value` is a numeric vector. We may do that like
so:

```{r}
conversion <- function(value, conversion_factor) {
  stopifnot(as.numeric(value))
  dat <- value
  converted <- dat*conversion_factor
  return(converted)
}
```

It still works when given proper input.

```{r}
conversion(value = 1, conversion_factor = 1000)
```

But fails instantly if given improper input.

```{r}
# value is a character instead of numeric
conversion(value = "one", conversion_factor = 1000)
conversion(value = "1", conversion_factor = 1000)
```


### Testing and Documenting

> ## Tip: Testing and documenting
>
> It's important to both test functions and document them:
> Documentation helps you, and others, understand what the
> purpose of your function is, and how to use it, and it's
> important to make sure that your function actually does
> what you think.
>
> When you first start out, your workflow will probably look a lot
> like this:
>
>  1. Write a function
>  2. Comment parts of the function to document its behaviour
>  3. Load in the source file
>  4. Experiment with it in the console to make sure it behaves
>     as you expect
>  5. Make any necessary bug fixes
>  6. Rinse and repeat.
>
> Formal documentation for functions, written in separate `.Rd`
> files, gets turned into the documentation you see in help
> files. The [roxygen2][] package allows R coders to write documentation
> alongside the function code and then process it into the appropriate `.Rd`
> files. You will want to switch to this more formal method of writing
> documentation when you start writing more complicated R projects.
>
> Formal automated tests can be written using the [testthat][] package.


[roxygen2]: http://cran.r-project.org/web/packages/roxygen2/vignettes/rd.html
[testthat]: http://r-pkgs.had.co.nz/tests.html







> ## Tip
>
> R has some unique aspects that can be exploited when performing more
> complicated operations. We have not written anything that requires
> knowledge of these more advanced concepts. In the future when you are
> comfortable writing functions in R, you can learn more by reading the
> [R Language Manual][man] or this [chapter][] from
> [Advanced R Programming][adv-r] by Hadley Wickham.

[man]: http://cran.r-project.org/doc/manuals/r-release/R-lang.html#Environment-objects
[chapter]: http://adv-r.had.co.nz/Environments.html
[adv-r]: http://adv-r.had.co.nz/
